---
title: "format spatial GEX data (semla)"
author: "Victoria Muliadi"
date: "2024-04-12"
output: html_document
---

Preprocess spatial GEX data and create Seurat object using 'semla' package (https://ludvigla.github.io/semla/)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r load packages}
library(semla)
library(readr)
```

## Load spatial GEX data as Seurat object

```{r load spatial gex data}

root.path <- "C:/Users/victo/Downloads/BC_spaceranger/"

visium_id <- "V19T26-032_A1"
tumor_area <- "BC-SA3-D"
section <- "D2"

feature.matrix.path <- "/filtered_feature_bc_matrix.h5"
spotfile.path <- "/spatial/tissue_positions_list.csv"
img.path <- "/spatial/tissue_hires_image.png"
json.path <- "/spatial/scalefactors_json.json"

samples <- paste0(root.path, visium_id, feature.matrix.path)
imgs <- paste0(root.path, visium_id, img.path)
spotfiles <- paste0(root.path, visium_id, spotfile.path)
json <- paste0(root.path, visium_id, json.path)

infoTable <- tibble(samples, imgs, spotfiles, json,
                    visium_id, tumor_area, section)

se <- ReadVisiumData(infoTable)

```

## Subset data to remove genes and spots with low expression

Criteria used in this example:
- each spot should have at least 500 UMI counts
- each gene should have at least 100 UMI counts and be expressed in at least 5 spots in the dataset
- remove mitochondrial and ribosomal genes + MALAT1 (long non-coding RNA)

```{r subset genes and spots}
# nCount: UMI (for columns / spots)
# nFeature: gene (for columns / spots)

## criteria:
# spots (columns): >= 500 UMI counts

keep.spots <- colnames(se)[se$nCount_Spatial >= 500]

# genes (rows): >= 100 UMI counts + expressed in >= 5 spots in dataset
# total expression / UMI count for the gene across all spots:
nCount.UMI <- Matrix::rowSums(GetAssayData(se))

# number of spots where gene is observed:
nObs.genes <- Matrix::rowSums(GetAssayData(se) > 0)

keep.genes <- rownames(se)[nCount.UMI >= 100 & nObs.genes >= 5]

## filter mitochondrial and ribosomal genes + MALAT1 (lncRNA)
keep.genes <- keep.genes[!grepl("MALAT1", keep.genes)]
keep.genes <- keep.genes[!grepl("^(MT-|RPL|RPS)", keep.genes)]

# subset
se <- SubsetSTData(se, spots = keep.spots, features = keep.genes)
```


## Load H&E images

```{r load H&E images}
se <- LoadImages(se)

ImagePlot(se)

```

## Save Seurat object as RDS file

```{r save seurat object}
write_rds(se, "C:/Users/victo/Downloads/spatialVDJ/scripts/sample_analysis/BCSA3_tumD.rds")

```

## Session information

```{r session info}
sessionInfo()
```


