---
title: "format clone count matrices (MiXCR)"
author: "Victoria Muliadi"
date: "2024-04-10"
output: html_document
---

Format clone count matrices (called by MiXCR) so they can be added as an Assay object to the Seurat object containing spatial GEX data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r load packages}
library(semla)
library(tidyverse)
library(Biostrings)
library(readr)
library(data.table)
```

## Load spatial GEX data (which has been subsetted to remove spots and genes with low expression)

```{r load spatial gex data}
se <- read_rds("C:/Users/victo/Downloads/spatialVDJ/scripts/sample_analysis/BCSA3_tumD.rds")

```

## Clone count matrices

Load clone descriptions (i.e. file containing VDJ calls, CDR3 sequences, etc.)
```{r load clone desc}

clones.main.dir <- list.files(path = "C:/Users/victo/Downloads/spatialVDJ/clones/MiXCR/BCSA3",
                              pattern = "clones.airr.tsv",
                              full.names = TRUE)

all.clones <- read_tsv(clones.main.dir)

head(all.clones)

```

```{r format clone description list}

# create new dataframe with chain names from clone description file
clone.names.cdr3 <- all.clones[, c("clone_id", "c_call", "v_call", "j_call", "cdr3", "cdr3_aa", "junction", "junction_aa")]
clone.names.cdr3$chain <- str_split_i(clone.names.cdr3$c_call, "C", 1)
clone.names.cdr3$chain <- str_split_i(clone.names.cdr3$chain, "[1234]", 1)
clone.names.cdr3$chain <- str_split_i(clone.names.cdr3$chain, "\\*", 1)

# sanity check if V call and J call are the same
sum(str_sub(clone.names.cdr3$v_call,1,3)!=str_sub(clone.names.cdr3$j_call,1,3))

# replace NA values in clone.names with J calls, else keep the existing chain name
clone.names.cdr3.new <- clone.names.cdr3 %>% mutate(chain = ifelse(is.na(chain), str_split_i(clone.names.cdr3$j_call, "J", 1), chain))

# check list of chains
table(clone.names.cdr3.new$chain)

# new clone name: chain + clone id
clone.names.cdr3.new$clone_name <- paste0(clone.names.cdr3.new$chain, "clone", clone.names.cdr3.new$clone_id)

# save clone list
write_csv(clone.names.cdr3.new, "C:/Users/victo/Downloads/spatialVDJ/scripts/sample_analysis/BCSA3_MiXCR_clones.csv")

# load clone list
clone.names.cdr3.new <- read_csv("C:/Users/victo/Downloads/spatialVDJ/scripts/sample_analysis/BCSA3_MiXCR_clones.csv")

```

## Load clonotype count matrices for each region

```{r load clone matrix}

# full.names: include full file paths
# recursive: look for files in subfolders
file.dir <- list.files(path = "C:/Users/victo/Downloads/spatialVDJ/clones/MiXCR/BCSA3",
           pattern = "UMIcount.tsv",
           full.names = TRUE, recursive = TRUE)

# load clone matrix for region D (sections 1 and 2) in this sample
clones.d1 <- read_tsv(file.dir[1])
clones.d2 <- read_tsv(file.dir[2])

# other regions present in this sample:
# clones.a2 <- read_tsv(file.dir[3])
# clones.c1 <- read_tsv(file.dir[4])
# clones.c2 <- read_tsv(file.dir[5])

```


## Format clone matrices

How-to:
- reverse complement spatial barcodes in count matrix

- add suffix to spatial barcodes
** NOTE: if adding >1 region to Seurat object, see note on INTEGRATION below

- transpose matrix (clone ID as rownames, spatial barcodes as colnames)

- filter clones with ID "-1" (i.e. not assigned to any clones)

- rename clone IDs with clone names based on clone description table

- remove spatial barcodes in clone matrix that are not present in Seurat object (e.g. spots that do not cover tissue, spots with low read counts etc)

- add spatial barcodes that are present in Seurat object but not in clone matrix and fill with zero
(before adding the matrix as a new Assay to the Seurat object, Seurat expects the new Assay to have the same number of spots as the original Seurat object)

** INTEGRATION **
if adding >1 region/section to Seurat object, suffixes must be added according to the order that the regions were added in the Seurat object
e.g. region 1 gets suffix "-1", region 2 gets "-2", etc
then join all the matrices together before adding the combined matrix as a new Assay

- before adding the clone matrix as an Assay, it could be a good idea to save this matrix as e.g. a delimited file (.csv)
NOTE!!! if you are trying to save files as .csv, there are 2 different functions that work slightly differently:
(1) write.csv() (utils package)
(2) write_csv() (readr package) -> does NOT save rownames in output file

since the rownames in this case contain the clone names, always be sure to save them in the output file!!! (or convert rownames to column before saving)

- add clone matrix as an Assay to Seurat object

```{r format clones section 1}

# rename first column to barcode
clones.d1.renamed <- clones.d1 %>% dplyr::rename(barcode = ...1)

# reverse complement barcodes
clones.d1.renamed$barcode <- as.vector(reverseComplement(DNAStringSet(clones.d1.renamed$barcode)))

# check if reverse complement barcodes in clone matrices are in spatial GEX data
paste0(clones.d1.renamed$barcode,"-1") %in% colnames(se.d1)

# add suffix to barcode
clones.d1.suffix <- clones.d1.renamed %>% mutate(barcode = paste0(barcode, "-1"))

# use spatial barcodes as rownames
clones.d1.t <- clones.d1.suffix %>% column_to_rownames(var = "barcode")

# transpose matrix
clones.d1.trans <- as.data.frame(t(clones.d1.t))

# rownames to column so clone_id can be renamed
clones.d1.trans.x <- clones.d1.trans %>% rownames_to_column("clone_id")

# filter clone_id -1 (not assigned to any clone)
clones.d1.trans.f <- clones.d1.trans.x %>%
  filter(clone_id!="-1")

# rename count matrix rows with new clone names
# clones that are present in matrix
clones.d1.matrix <- clones.d1.trans.f
clones.d1.matrix$clone_id <- as.numeric(clones.d1.matrix$clone_id)
clones.d1.matrix <- clones.d1.matrix %>% arrange(clone_id)

# add new clone names
# left join (similar to cbind? but no restrictions on dimensions, requires same column name in both, left join means that the structure of the df on the left side will not be changed)
clones.d1.matrix.joined <- left_join(clones.d1.matrix, clone.names.cdr3.new[, c("clone_id", "clone_name")], by = "clone_id")
clones.d1.matrix.joined <- clones.d1.matrix.joined %>% column_to_rownames("clone_name")
clones.d1.matrix.joined <- clones.d1.matrix.joined[, -1]

# get barcodes present in spatial GEX data
gex.bc.d1 <- colnames(se.d1)

# remove spatial barcodes present in count matrix but not present in seurat object
clones.d1.matrix.filtered <- clones.d1.matrix.joined[, colnames(clones.d1.matrix.joined) %in% gex.bc.d1]

# add spatial barcodes present in seurat object but not present in count matrix as new column and fill with zeros
# create vector of barcodes missing from clone matrix
diff.bc.d1 <- setdiff(gex.bc.d1, colnames(clones.d1.matrix.filtered))

# create dataframe filled with 0, same number of rows as count matrix, same number of columns as diff.bc
df.d1 <- data.frame(matrix(0, nrow = nrow(clones.d1.matrix.filtered), ncol=length(diff.bc.d1)))

# set row names and column names of df to match the clone names of count matrix (row) and missing barcodes (column)
rownames(df.d1) <- rownames(clones.d1.matrix.filtered)
colnames(df.d1) <- diff.bc.d1

# join count matrix and df by columns
clones.d1.matrix.filtered <- cbind(clones.d1.matrix.filtered, df.d1)

# save clone matrix
write.csv(clones.d1.matrix.filtered, "C:/Users/victo/Downloads/spatialVDJ/scripts/sample_analysis/BCSA3_D1_clone_matrix_mixcr.csv")

# change rowname to common column name for integration
int.clones.d1.matrix.filtered <- clones.d1.matrix.filtered %>% rownames_to_column(var = "clone_name")

```


```{r format clones section 2}

# rename first column to barcode
clones.d2.renamed <- clones.d2 %>% dplyr::rename(barcode = ...1)

# reverse complement barcodes
clones.d2.renamed$barcode <- as.vector(reverseComplement(DNAStringSet(clones.d2.renamed$barcode)))

# check if reverse complement barcodes in clone matrices are in spatial GEX data
paste0(clones.d2.renamed$barcode,"-1") %in% colnames(se.d2)

# add suffix to barcode
clones.d2.suffix <- clones.d2.renamed %>% mutate(barcode = paste0(barcode, "-1"))

# use spatial barcodes as rownames
clones.d2.t <- clones.d2.suffix %>% column_to_rownames(var = "barcode")

# transpose matrix
clones.d2.trans <- as.data.frame(t(clones.d2.t))

# rownames to column so clone_id can be renamed
clones.d2.trans.x <- clones.d2.trans %>% rownames_to_column("clone_id")

# filter clone_id -1 (not assigned to any clone)
clones.d2.trans.f <- clones.d2.trans.x %>%
  filter(clone_id!="-1")

# rename count matrix rows with new clone names
# clones that are present in matrix
clones.d2.matrix <- clones.d2.trans.f
clones.d2.matrix$clone_id <- as.numeric(clones.d2.matrix$clone_id)
clones.d2.matrix <- clones.d2.matrix %>% arrange(clone_id)

# add new clone names
# left join (similar to cbind? but no restrictions on dimensions, requires same column name in both, left join means that the structure of the df on the left side will not be changed)
clones.d2.matrix.joined <- left_join(clones.d2.matrix, clone.names.cdr3.new[, c("clone_id", "clone_name")], by = "clone_id")
clones.d2.matrix.joined <- clones.d2.matrix.joined %>% column_to_rownames("clone_name")
clones.d2.matrix.joined <- clones.d2.matrix.joined[, -1]

# get barcodes present in spatial GEX data
gex.bc.d2 <- colnames(se.d2)

# remove spatial barcodes present in count matrix but not present in seurat object
clones.d2.matrix.filtered <- clones.d2.matrix.joined[, colnames(clones.d2.matrix.joined) %in% gex.bc.d2]

# add spatial barcodes present in seurat object but not present in count matrix as new column and fill with zeros
# create vector of barcodes missing from clone matrix
diff.bc.d2 <- setdiff(gex.bc.d2, colnames(clones.d2.matrix.filtered))

# create dataframe filled with 0, same number of rows as count matrix, same number of columns as diff.bc
df.d2 <- data.frame(matrix(0, nrow = nrow(clones.d2.matrix.filtered), ncol=length(diff.bc.d2)))

# set row names and column names of df to match the clone names of count matrix (row) and missing barcodes (column)
rownames(df.d2) <- rownames(clones.d2.matrix.filtered)
colnames(df.d2) <- diff.bc.d2

# join count matrix and df by columns
clones.d2.matrix.filtered <- cbind(clones.d2.matrix.filtered, df.d2)

# save clone matrix
write.csv(clones.d2.matrix.filtered, "C:/Users/victo/Downloads/spatialVDJ/scripts/sample_analysis/BCSA3_D2_clone_matrix_mixcr.csv")

```


```{r format clone matrix for integration section 2}

# change barcode suffixes
int.clones.d2.matrix.filtered <- clones.d2.matrix.filtered

setnames(int.clones.d2.matrix.filtered, old = colnames(int.clones.d2.matrix.filtered), new = paste0(str_split_i(colnames(int.clones.d2.matrix.filtered), pattern = "-", 1), "-2"))

# change rowname to common column name for integration
int.clones.d2.matrix.filtered <- int.clones.d2.matrix.filtered %>% rownames_to_column(var = "clone_name")

```

## Integration

NOTE: number of spots in GEX data should be equal to number of spots in combined clone matrix!!!

how-to:
- full_join the matrices (that were formatted in previous steps) one by one
- replace NA with zero (otherwise Seurat will throw an error)

```{r merge clone matrices of all sections}
# join matrices together
combined.clones <- full_join(int.clones.d1.matrix.filtered, int.clones.d2.matrix.filtered, by = "clone_name")

# change clone_name column back to rowname
combined.clones.new <- combined.clones %>% column_to_rownames(var = "clone_name")

# replace NA with zero
combined.clones.new <- combined.clones.new %>% mutate_all(~ replace(., is.na(.), 0))

# alternative:
# combined.clones.new[is.na(combined.clones.new)] <- 0

# save merged clone matrix
write.csv(combined.clones.new, "C:/Users/victo/Downloads/spatialVDJ/scripts/sample_analysis/BCSA3_tumD_clone_matrix_mixcr.csv")
# write_tsv(combined.clones.new,"C:/Users/victo/Downloads/spatialVDJ/scripts/sample_analysis/BCSA3_tumD_clone_matrix_mixcr.tsv")

# create new assay in seurat object
se[["Clone"]] <- CreateAssayObject(combined.clones.new)

# save seurat object
write_rds(se, "C:/Users/victo/Downloads/spatialVDJ/scripts/sample_analysis/BCSA3_tumD_clones_mixcr.rds")

```

## Session information

```{r session info}
sessionInfo()
```


